---
title: "Microbial Metagenomics Portfolio 2023 — Project Story"
subtitle: "From Raw Reads to Metagenome-Assembled Genomes: A Soil Metagenomics Pipeline"
author: "Skylar Zeng"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
    fig_width: 9
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  eval       = FALSE,   # set to TRUE when input data are present
  fig.align  = "center"
)
```

---

# Overview

This document narrates the complete computational pipeline used in my undergraduate
thesis on soil microbial community structure across a land-use gradient.
It is structured as a technical portfolio artifact: every processing step is
described, the key code is embedded verbatim, and representative results are
discussed. Raw sequencing data are not included; code logic and workflow steps
are fully preserved.

**Biological Question**

> How does microbial community composition and functional potential vary across a
> forest-to-agriculture disturbance gradient, and can metagenome-assembled genomes
> (MAGs) recovered from these soils reveal novel lineages involved in nitrogen and
> carbon cycling?

**Study design**: 24 soil samples (12 pristine forest, 12 agricultural) collected
along a single land-use transect. Each sample was subjected to whole-shotgun
metagenomic sequencing (Illumina NovaSeq 6000, 2 × 150 bp, ~10 Gb per sample).

---

# Environment Setup

All compute-intensive steps ran on a SLURM HPC cluster. The Conda environment
below reproduces the analysis locally.

```{r env-setup, eval=FALSE}
# Create and activate the Conda environment
# (run in terminal, not R)
# conda env create -f docs/environment.yml
# conda activate metagenomics-2023

# R packages
required_pkgs <- c(
  "tidyverse", "vegan", "phyloseq", "ggplot2", "ggrepel",
  "ggpubr", "rstatix", "DESeq2", "ComplexHeatmap", "circlize",
  "pheatmap", "RColorBrewer", "patchwork"
)
missing_pkgs <- required_pkgs[!required_pkgs %in% installed.packages()[, "Package"]]
if (length(missing_pkgs) > 0) {
  BiocManager::install(missing_pkgs)
}
```

---

# Pipeline Steps

## Step 1: Quality Control

FastQC assessed raw read quality; Trimmomatic removed adapters and low-quality
bases. SLURM array jobs processed all 24 samples in parallel.

```{bash fastqc-slurm, eval=FALSE, engine="bash"}
#!/bin/bash
#SBATCH --job-name=fastqc_raw
#SBATCH --array=1-24
#SBATCH --cpus-per-task=4
#SBATCH --mem=8G
#SBATCH --time=02:00:00

module load fastqc/0.11.9

SAMPLE=$(sed -n "${SLURM_ARRAY_TASK_ID}p" sample_list.txt)
fastqc --threads 4 --outdir qc/fastqc_raw \
    raw_reads/${SAMPLE}_R1.fastq.gz \
    raw_reads/${SAMPLE}_R2.fastq.gz
```

```{bash trimmomatic-slurm, eval=FALSE, engine="bash"}
#!/bin/bash
#SBATCH --job-name=trimmomatic
#SBATCH --array=1-24
#SBATCH --cpus-per-task=8
#SBATCH --mem=16G
#SBATCH --time=04:00:00

module load trimmomatic/0.39

SAMPLE=$(sed -n "${SLURM_ARRAY_TASK_ID}p" sample_list.txt)

trimmomatic PE -threads 8 -phred33 \
    raw_reads/${SAMPLE}_R1.fastq.gz \
    raw_reads/${SAMPLE}_R2.fastq.gz \
    trimmed/${SAMPLE}_R1_paired.fastq.gz   trimmed/${SAMPLE}_R1_unpaired.fastq.gz \
    trimmed/${SAMPLE}_R2_paired.fastq.gz   trimmed/${SAMPLE}_R2_unpaired.fastq.gz \
    ILLUMINACLIP:TruSeq3-PE-2.fa:2:30:10:8:true \
    LEADING:3 TRAILING:3 SLIDINGWINDOW:4:20 MINLEN:50
```

**Result**: Average read retention 94.8% across 24 samples. MultiQC confirmed
all samples passed quality thresholds (Q30 > 90%).

---

## Step 2: Host Decontamination

Plant host reads were removed by mapping against the *Zea mays* B73 reference
genome with Bowtie2. Only unmapped (microbial) read pairs were retained.

```{bash host-decontam, eval=FALSE, engine="bash"}
#!/bin/bash
#SBATCH --cpus-per-task=16
#SBATCH --mem=32G

SAMPLE=$(sed -n "${SLURM_ARRAY_TASK_ID}p" sample_list.txt)

bowtie2 -x databases/host_bt2_index \
    -1 trimmed/${SAMPLE}_R1_paired.fastq.gz \
    -2 trimmed/${SAMPLE}_R2_paired.fastq.gz \
    --threads 16 --sensitive \
    --un-conc-gz decontaminated/${SAMPLE}_clean_R%.fastq.gz \
    -S /dev/null \
    2> decontaminated/${SAMPLE}_bowtie2_host.log
```

**Result**: Mean 2.3% host mapping rate. Retained ~97.7% of reads for
downstream microbial analysis.

---

## Step 3: Taxonomic Profiling

### Kraken2 k-mer Classification

```{bash kraken2, eval=FALSE, engine="bash"}
#!/bin/bash
#SBATCH --cpus-per-task=16
#SBATCH --mem=128G
#SBATCH --partition=highmem

SAMPLE=$(sed -n "${SLURM_ARRAY_TASK_ID}p" sample_list.txt)

kraken2 \
    --db   databases/kraken2_standard_20230314 \
    --threads 16 --confidence 0.1 \
    --paired --gzip-compressed \
    --output  kraken2/${SAMPLE}.kraken2 \
    --report  kraken2/${SAMPLE}.k2report \
    decontaminated/${SAMPLE}_R1.fastq.gz \
    decontaminated/${SAMPLE}_R2.fastq.gz
```

### Bracken Abundance Re-estimation

```{bash bracken, eval=FALSE, engine="bash"}
bracken -d databases/kraken2_standard_20230314 \
    -i  kraken2/${SAMPLE}.k2report \
    -o  bracken/species/${SAMPLE}_bracken_species.txt \
    -r  150 -l S -t 10
```

### Parsing Kraken2 Reports (Python)

```{python parse-kraken2, eval=FALSE}
# scripts/python/parse_kraken2_reports.py (abbreviated)
import pandas as pd
from pathlib import Path

K2_COLS = ["pct", "clade_reads", "taxon_reads", "rank", "taxid", "name"]

def parse_k2_report(path: Path, rank: str) -> pd.Series:
    df = pd.read_csv(path, sep="\t", header=None, names=K2_COLS)
    df["name"] = df["name"].str.strip()
    subset = df[df["rank"] == rank].set_index("name")["taxon_reads"]
    return subset

# Aggregate all samples into one matrix
report_files = sorted(Path("kraken2").glob("*.k2report"))
combined = pd.DataFrame({f.stem: parse_k2_report(f, "S") for f in report_files})
combined.fillna(0).astype(int).to_csv("data/kraken2_species_matrix.tsv", sep="\t")
```

**Result**: Mean classification rate 68% (range 52–79%), consistent with
high novelty in forest soils relative to the reference database.

---

## Step 4: De Novo Metagenomic Assembly

MEGAHIT was used for per-sample assembly. Co-assembly was rejected to preserve
strain-level diversity across the environmental gradient.

```{bash megahit, eval=FALSE, engine="bash"}
#!/bin/bash
#SBATCH --cpus-per-task=24
#SBATCH --mem=128G
#SBATCH --time=24:00:00

SAMPLE=$(sed -n "${SLURM_ARRAY_TASK_ID}p" sample_list.txt)

megahit \
    -1 decontaminated/${SAMPLE}_R1.fastq.gz \
    -2 decontaminated/${SAMPLE}_R2.fastq.gz \
    --out-dir     assembly/${SAMPLE} \
    --out-prefix  ${SAMPLE} \
    --num-cpu-threads 24 \
    --k-list 21,29,39,59,79,99,119,141 \
    --min-contig-len 1000
```

**Result**: Mean N50 of 2,841 bp (forest) and 2,284 bp (agricultural);
larger contigs in forest samples consistent with more coherent genomes
in higher-diversity communities.

---

## Step 5: Read Mapping for Binning Coverage

```{bash bowtie2-mapping, eval=FALSE, engine="bash"}
bowtie2-build --threads 16 assembly/${SAMPLE}/${SAMPLE}.contigs.fa mapping/${SAMPLE}_index

bowtie2 -x mapping/${SAMPLE}_index \
    -1 decontaminated/${SAMPLE}_R1.fastq.gz \
    -2 decontaminated/${SAMPLE}_R2.fastq.gz \
    --threads 16 --no-unal -S mapping/${SAMPLE}.sam \
    2> mapping/${SAMPLE}_mapping.log

samtools view -bS -@ 16 mapping/${SAMPLE}.sam | \
samtools sort    -@ 16 -o mapping/${SAMPLE}_sorted.bam
samtools index mapping/${SAMPLE}_sorted.bam
rm mapping/${SAMPLE}.sam   # remove large SAM
```

---

## Step 6: Genome Binning with MetaBAT2

```{bash metabat2, eval=FALSE, engine="bash"}
# Compute per-contig coverage depth
jgi_summarize_bam_contig_depths \
    --outputDepth bins/metabat2/${SAMPLE}_depth.txt \
    mapping/${SAMPLE}_sorted.bam

# Bin contigs
metabat2 \
    -i assembly/${SAMPLE}/${SAMPLE}.contigs.fa \
    -a bins/metabat2/${SAMPLE}_depth.txt \
    -o bins/metabat2/${SAMPLE}/${SAMPLE}_bin \
    -m 2500 -t 16 --unbinned --saveCls
```

**Result**: 108 total bins recovered across 24 samples before quality filtering.

---

## Step 7: MAG Quality Assessment (CheckM2)

```{bash checkm2, eval=FALSE, engine="bash"}
checkm2 predict \
    --input bins/metabat2/${SAMPLE} \
    --output-directory checkm2/${SAMPLE} \
    --extension fa \
    --database-path databases/checkm2_db/uniref100.KO.1.dmnd \
    --threads 16 --force
```

### Summarising CheckM2 Results (Python)

```{python summarize-checkm, eval=FALSE}
# scripts/python/summarize_checkm_results.py (abbreviated)
import pandas as pd
from pathlib import Path

HQ_COMP, HQ_CONT = 90.0, 5.0
MQ_COMP, MQ_CONT = 70.0, 10.0

def assign_tier(row):
    if row.Completeness >= HQ_COMP and row.Contamination <= HQ_CONT:
        return "high"
    if row.Completeness >= MQ_COMP and row.Contamination <= MQ_CONT:
        return "medium"
    return "low"

reports = sorted(Path("checkm2").glob("*/quality_report.tsv"))
combined = pd.concat(
    [pd.read_csv(r, sep="\t").assign(sample_id=r.parent.name) for r in reports]
)
combined["quality_tier"] = combined.apply(assign_tier, axis=1)
combined.to_csv("data/checkm2_all_samples_summary.csv", index=False)

print(combined["quality_tier"].value_counts())
```

**Result**: 47 MAGs passed QC (12 high-quality, 35 medium-quality).

---

## Step 8: Phylogenomic Classification (GTDB-Tk)

```{bash gtdbtk, eval=FALSE, engine="bash"}
# Run on all passing MAGs together
gtdbtk classify_wf \
    --genome_dir bins/passing \
    --out_dir    gtdbtk \
    --extension  fa \
    --cpus       48 \
    --skip_ani_screen
```

**Result**: 3 MAGs placed in unnamed Acidobacteriota families; 2 in candidate
phylum FCPU426 — entirely novel to this soil type.

---

## Step 9: Alpha-Diversity Analysis

```{r alpha-diversity}
library(tidyverse)
library(vegan)
library(ggpubr)
library(rstatix)

# Load species-level Bracken matrix (taxa × samples)
otu_wide <- read_tsv("data/bracken_species_matrix.tsv") %>%
  column_to_rownames("taxon")
otu_mat <- t(as.matrix(otu_wide))

# Rarefy to minimum depth
set.seed(42)
min_reads <- min(rowSums(otu_mat))
otu_rarefied <- rrarefy(otu_mat, sample = min_reads)

# Compute indices
alpha_df <- data.frame(
  sample_id = rownames(otu_rarefied),
  shannon   = diversity(otu_rarefied, index = "shannon"),
  simpson   = diversity(otu_rarefied, index = "simpson"),
  richness  = specnumber(otu_rarefied)
) %>%
  mutate(evenness = shannon / log(richness)) %>%
  left_join(read_csv("data/metadata.csv"), by = "sample_id")

# Wilcoxon tests
stats_results <- alpha_df %>%
  pivot_longer(cols = c(shannon, simpson, richness, evenness)) %>%
  group_by(name) %>%
  wilcox_test(value ~ land_use) %>%
  adjust_pvalue("BH") %>%
  add_significance("p.adj")

print(stats_results)
```

**Key result**: Shannon H' significantly higher in forest soils
(Wilcoxon, BH-adjusted *p* < 0.01), consistent with disturbance-driven
diversity loss.

---

## Step 10: Beta-Diversity and Community Ordination

```{r beta-diversity}
library(vegan)

bc_dist <- vegdist(otu_rarefied, method = "bray")

# PERMANOVA
set.seed(42)
permanova <- adonis2(bc_dist ~ land_use, data = metadata,
                     permutations = 9999, by = "margin")
print(permanova)
# R² = 0.42, p = 0.001

# PCoA
pcoa <- cmdscale(bc_dist, eig = TRUE, k = 2)
pct_var <- round(pcoa$eig / sum(pcoa$eig) * 100, 1)

pcoa_df <- data.frame(
  sample_id = rownames(pcoa$points),
  PC1 = pcoa$points[, 1],
  PC2 = pcoa$points[, 2]
) %>% left_join(metadata, by = "sample_id")

ggplot(pcoa_df, aes(PC1, PC2, colour = land_use)) +
  geom_point(size = 3) +
  stat_ellipse(level = 0.95) +
  labs(x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)")) +
  scale_colour_manual(values = c(forest = "#2d6a4f", agricultural = "#b5838d")) +
  theme_bw()
```

**Key result**: PERMANOVA R² = 0.42, *p* = 0.001 — land use explains 42% of
community composition variance.

---

## Step 11: Differential Abundance (DESeq2)

```{r deseq2}
library(DESeq2)

# Integer genus-level count matrix (taxa × samples)
count_mat <- read_tsv("data/bracken_genus_matrix.tsv") %>%
  column_to_rownames("taxon") %>%
  as.matrix() %>% round()

dds <- DESeqDataSetFromMatrix(
  countData = count_mat[, metadata$sample_id],
  colData   = metadata,
  design    = ~ land_use
)

# Low-count filter: keep genera with ≥ 10 reads in ≥ 3 samples
dds <- dds[rowSums(counts(dds) >= 10) >= 3, ]
dds <- DESeq(dds, quiet = TRUE)

res <- results(dds,
               contrast = c("land_use", "agricultural", "forest"),
               alpha = 0.05, pAdjustMethod = "BH")

# Significant genera (|log2FC| > 1, padj < 0.05)
sig <- as.data.frame(res) %>%
  rownames_to_column("taxon") %>%
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) > 1) %>%
  arrange(padj)

nrow(sig)   # number of differentially abundant genera
```

---

## Step 12: Functional Gene Heatmap

```{r functional-heatmap}
library(ComplexHeatmap)

# Presence/absence of key KEGG Orthologs in each MAG
# (full code in scripts/r/04_mag_functional_heatmap.R)

# Target genes for nitrogen and carbon cycling
target_ko <- c(
  nifH = "K02588", nifD = "K02586",     # N fixation
  nosZ = "K00376", norB = "K04561",     # Denitrification
  amoA = "K10944",                       # Nitrification
  mcrA = "K00399",                       # Methanogenesis
  rbcL = "K01601"                        # C fixation (CBB)
)

# pa_mat: rows = MAGs, columns = KOs (0/1)
# (populated from eggNOG-mapper *.annotations files)
Heatmap(
  pa_mat,
  name   = "Presence",
  col    = c("0" = "#f0f0f0", "1" = "#1d3557"),
  column_split = gene_pathways,
  cluster_rows = FALSE, cluster_columns = FALSE
)
```

**Key result**: *nifH* (nitrogen fixation) and *nosZ* (denitrification) genes
enriched in forest-associated MAGs; agricultural MAGs carried more
*amoA* (nitrification) genes — consistent with fertiliser-driven nitrifier
enrichment.

---

# Results Summary

| Analysis | Key Metric | Forest | Agricultural | Significance |
|---|---|---|---|---|
| α-diversity (Shannon H') | mean ± SD | 4.12 ± 0.31 | 3.47 ± 0.42 | *p* < 0.01 |
| Species richness | mean | 1 842 | 1 203 | *p* < 0.01 |
| β-diversity (PERMANOVA) | R² | — | — | 0.42, *p* = 0.001 |
| MAGs recovered | count | 31 | 16 | — |
| High-quality MAGs | count | 9 | 3 | — |
| Novel Acidobacteria | unnamed families | 3 | 0 | — |
| *nifH*-carrying MAGs | count | 7 | 1 | — |

---

# Reproducibility Notes

```{r session-info}
sessionInfo()
```

All SLURM scripts embed exact module versions used.  
All R scripts conclude with `sessionInfo()`.  
Conda environment: `docs/environment.yml`.  
Random seeds set to `42` throughout.

---

*Undergraduate thesis project, 2023. Code logic and workflow steps preserved
as a public technical portfolio artifact.*
